Prospects of detecting gamma-ray emission from galaxy clusters: cosmic rays and dark matter annihilations

### 2011-08-03 paper.tex
removed C and del

In this paper, we will investigate in some more detail the potential
of several of the most promising galaxy clusters to produce an
annihilation gamma-ray yield which could be observable with present
and planned gamma-ray detectors. For convenience, we summarize in
Table~\ref{tab:flux_tab} our results for the gamma-ray flux from four
promising clusters with a high gamma-ray yield. We will in the
following derive the emission from two different DM models and a
hadronic CR model to show that Virgo has the highest gamma-ray flux
from annihilating DM. \C{However, if the boost from Sommerfeld
  enhancement (SFE) is small then the gamma-ray emission induced by
  CRs is dominating most clusters with the exception of a few
  clusters; most notably Fornax that shows the largest ratio of DM
  annihilation-to-CR flux. In particular, we identify Fornax and Virgo
  as prime targets for DM observations,} Perseus for CR induced
emission and we additionally decide in favor of the well studied
cluster Coma for comparison. While studying these four clusters in
great detail, we complement the analysis by calculating the expected
gamma-ray fluxes of a nearby X-ray flux complete sample of galaxy
clusters from the Rosat all-sky survey (extended HIghest X-ray FLUx
Galaxy Cluster Sample, HIFLUGCS,
\citealt{2002ApJ...567..716R,2007A&A...466..805C}).\footnote{Note that
  we have added the Virgo cluster to the sample that we nevertheless
  refer to as extended HIFLUGCS catalogue in the following.} For
previous work related to DM in clusters, see, e.g.,
\cite{2006A&A...455...21C,2009PhRvD..80b3005J,2011arXiv1104.3530S,2011ApJ...726L...6C,2010JCAP...05..025A}
and for recent work on CRs in clusters see
\cite{2010MNRAS.409..449P,2011MNRAS.410..127B,2008MNRAS.385.1211P,2009JCAP...08..002K,2010MNRAS.407.1565D}.

\begin{table*}
\begin{minipage}{2.0\columnwidth}
  \caption{Gamma-ray flux from various clusters within $\rvir$.}
\begin{tabular}{l c c c c c c c c c c}
\hline
\hline
 Cluster &
\multicolumn{3}{c}{$F_{\gamma}(>100\,\rmn{MeV})$ $[\rmn{ph}\,\rmn{cm}^{-2}\,\rmn{s}^{-1}]$:} & &
\multicolumn{3}{c}{$F_{\gamma}(>100\,\rmn{GeV})$ $[\rmn{ph}\,\rmn{cm}^{-2}\,\rmn{s}^{-1}]$:} &
$\mvir$ $^{(1)}$ & $\B_\rmn{sfe} $$^{(2)}$ & $\B_\rmn{sub} $$^{(3)}$ \\
         & DM-LP $^{(4)}$ & DM-BM-$\Kp$ $^{(5)}$ & CR-$\pi^0$ $^{(6)}$
         & & DM-LP $^{(4)}$ & DM-BM-$\Kp$ $^{(5)}$ & CR-$\pi^0$ $^{(6)}$ & $[10^{14}\,\msun]$ && \\
 \hline
 Coma & $8.2\times10^{-8}$ & $7.6\times10^{-11}$ & $4.1\times10^{-9}$
 & \,\,\,\,\, & $1.2\times10^{-11}$ & $2.8\times10^{-13}$ & $1.5\times10^{-12}$
 & $12.9$ & $530/65$ & $1290$ \\
 Perseus \,\,\,\,\,\, & $8.6\times10^{-8}$ & $7.8\times10^{-11}$ & $1.5\times10^{-8}$
 & \,\,\,\,\, & $1.2\times10^{-11}$ & $2.9\times10^{-13}$ & $5.5\times10^{-12}$
 & $8.6$ & $530/75$ & $1190$ \\
 Virgo & $1.6\times10^{-6}$ & $1.5\times10^{-9}$ & $1.5\times10^{-8}$
 & \,\,\,\,\, & $2.3\times10^{-10}$ & $5.4\times10^{-12}$ & $5.7\times10^{-12}$
 & $6.9$ & $530/80$ & $1120$ \\
 Fornax & $3.5\times10^{-7}$ & $3.2\times10^{-10}$ & $3.1\times10^{-10}$
 & \,\,\,\,\, & $5.1\times10^{-11}$ & $1.2\times10^{-12}$ & $1.1\times10^{-13}$
 & $2.4$ & $530/110$ & $890$ \\
\hline
\hline
\end{tabular}
\begin{quote}
  Notes: \\
  (1) The mass of Fornax, Coma, and Perseus are derived from \cite{2007A&A...466..805C},
  while the mass of Virgo is derived from \cite{1984ApJ...281...31T}.\\
  (2) The boost due to Sommerfeld enhancement. The first value shows the saturated
  Sommerfeld boost realized when substructures are present, the latter is the Sommerfeld
  boost without substructures (see Section II A 1).\\
  (3) \C{The boost due to substructures relative a the smooth DM
  distribution.} We integrate the emission in a cylinder with a radial
  extent of $2.5 R_{200}$ along the line-of-sight and an angular size
  corresponding to $\rvir$ (both measured from the cluster center).\\
  (4) The total gamma-ray flux from leptophilic (LP) DM where both the
  boost from substructures and Sommerfeld enhancement are
  included. Note that all these values are in conflict with upper limits from
  \Fermi (see the following sections for detail).\\
  (5) The total gamma-ray flux from the supersymmetric $\Kp$ benchmark (BM)
  model where the boost from substructures is included. See Sections II A 2 and II B 3.\\
  (6) Gamma-ray flux induced by CR protons. See Section V C.
 \label{tab:flux_tab}
  \end{quote}
\end{minipage}
\end{table*}

@@

We adopt our benchmark LP model from \cite{Finkbeiner:2010sm} where we
use a DM mass $m_\chi=1.6$~TeV and a branching ratio of
($1/4:1/4:1/2$) into ($\mu^+\mu^-:\e^+\e^-:\pi^+\pi^-$). \C{Furthermore,
since most of the gamma-ray flux is expected to come from dense
subhalos within clusters which have
a velocity dispersion} close to the velocity limit
$v_\rmn{sat}$ where the SFE saturates, we use
$S=\B_\sfe(v_\rmn{sat})\approx 530$ for the cluster
halos.\footnote{Note that $S$ is strictly speaking the Sommerfeld
  enhanced annihilation cross-section, but here we have additionally
  included the order unity constraint that ensures that the relic
  density is not over-depleted at freeze-out.} However, for figures
where we neglect the substructure boost, we adopt a velocity dependent
SFE that is normalized to fit the electron and positron excess at
Earth observed by the PAMELA and \Fermi satellites,
\begin{equation}
%\B_\sfe(\sigv) = 0.266\, \frac{c}{v_\rmn{sat}}\, \frac{v_\rmn{MW}}{\sigv} =
%121 \left(\frac{\mvir}{10^{15}\,\msun}\right)^{-1/3}\,.
\C{\B_\sfe(\sigv) = \B_\rmn{sfe,MW} \frac{v_\rmn{MW}}{\sigv} =
70 \left(\frac{\mvir}{10^{15}\,\msun}\right)^{-1/3}\,.}
\label{eq:B_sfe}
\end{equation}
\C{Here the boost factor needed to explain the PAMELA/\Fermi excess of
  electrons and positrons $\B_\rmn{sfe,MW}=300$, the velocity
  dispersion for the MW $v_\rmn{MW} \approx 220\,\rmn{km\,s}^{-1}$, and the
  velocity dispersion $\sigv$ for clusters is derived in
  \cite{2005RvMP...77..207V}.} Clusters have a larger velocity
dispersion and an accordingly smaller Sommerfeld boost factor.

@@

\subsubsection{Substructures}
\label{sect:subst}
High-resolution dissipationless DM simulations of MW type halos find
substantial amount of substructures in the periphery of DM halos,
while the substructures in the center suffer from dynamical friction
and tidal effects depleting their central number densities. Since the
DM annihilation rate depends on the density squared, the resulting
flux from substructures is boosted compared to the smooth density
distribution. While there is still a discrepancy in the literature of
the exact value of the predicted boost factor of the DM annihilation
luminosity from substructures in DM halos, this inconsistency starts
to become resolved, apparently converging toward predictions at the
high end \protect \cite{2008MNRAS.391.1685S, 2008Natur.456...73S,
  2010ApJ...718..899A}. \C{Following recent high-resolution
  simulations of Galaxy-sized halos, we adopt a boost factor due to
  substructures of 230 for such a halo \cite{2008MNRAS.391.1685S}
  (which needs to be scaled to cluster halos as we will discuss
  below).}

The initial suggestions of a small {\em total} boost ($<R_{200}$) of
order unity by the Via Lactea simulations \cite{2008JPhCS.125a2008K}
made assumptions of computing the boost assuming that the substructure
luminosity follows that of the smooth DM distribution. Physically,
this would imply substructure to follow the smooth DM distribution and
a radially independent concentration parameter of substructures. Both
assumptions are in conflict with tidal mass loss of satellites which
is at work in simulations and cause the radial number density to be
anti-biased with respect to the host's mass density profile yielding a
significantly flatter subhalo distribution compared to the smooth DM
distribution \cite{2010ApJ...718..899A}. This increases the
substructure boost preferentially in the outer parts of DM halos with
a factor that ranges between 20 and 1000 (for dwarfs and galaxy
clusters), hence depending sensitively on the host halo mass (see
Figure~\ref{fig:radial_lum}). \C{Recent {\it Phoenix} simulations by
  Gao et al. \cite{2011arXiv1107.1916G} confirm the large enhancement
  due to substructures in clusters. In fact they find a boost of 1125
  for a Coma like cluster which agrees within 10\% with our estimated
  boost (see e.g. Table~\ref{tab:flux_tab}).}

@@

In Fig.~\ref{fig:diff_BM} we show the differential flux from the
Fornax cluster which is one of the best clusters for indirect DM
searches due to its high DM annihilation fluxes and low CR induced
fluxes. We show the emission of four different supersymmetric BM
models and contrast it to the emission induced by CRs. Comparing this
emission to the differential gamma-ray upper limits set by \Fermi-LAT,
we find that the upper limits are not violated. The predicted DM flux
that is dominated by the continuum emission from $\Kp$ and $\Ip$
models (shown in the left panels) are about \C{a factor few} below the
upper limits, making it hard for \Fermi to probe these kind of DM
models \C{in the near future} without a significant improvement in the
analysis from e.g. stacking of clusters and improved methods for
analyzing extended sources. Furthermore, the gamma-ray signal induced
by CRs is expected to be about a factor 10 below the DM continuum
flux from the $\Kp$ and $\Ip$ BM models at 10~GeV. However, the IC
emission from upscattered CMB and SD photons is at least a factor of a
few lower for $\Kp$ and a factor 1000 lower for $\Ip,\Jp,\Js$, than
the expected flux from CRs above 100~MeV, making it very hard to
distinguish such a signal from the foreground due to the similar
spectral index. For energies below 100~MeV we expect IC emission by
primary shock accelerated electrons to be dominating
\cite{2010MNRAS.409..449P} over the supersymmetric DM induced
leptons. Hence in clusters, the IC emission from supersymmetric DM BM
models can be neglected compared to the CR pion and DM continuum
emission.

\begin{figure*}
\begin{minipage}{2.0\columnwidth}
 \includegraphics[width=0.49\columnwidth]{figures/flux.BMcompI.v14.0.1deg.1.6T.SubMass.IR2.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.BMcompJ.v14.0.1deg.1.6T.SubMass.IR2.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.BMcompK.v14.0.1deg.1.6T.SubMass.IR2.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.BMcompJs.v14.0.1deg.1.6T.SubMass.IR2.noMW.woGal.eps}
\caption{\it Comparing the differential flux from different models: we
  show the continuum emission from DM benchmark (BM) models (blue
  dashed), electrons and positrons from DM BM models that
  Compton-upscatter CMB photons (green dotted) as well as dust and
  star photons (orange dash-dotted), and CR induced gamma-ray emission
  (red solid). Each panel is associated with an individual DM BM
  model; upper left $\Ip$, lower left $\Kp$, upper right $\Jp$, and
  lower right $\Js$. The emission is calculated for the Fornax cluster
  using a point spread function of $0.1\deg$. The substructures boost
  the gamma-ray flux from IC upscattered CMB and continuum emission by
  a factor of 890 while the IC upscattered SD photons are only boosted
  by 20. Extended upper limits ($<1^\circ$) from \Fermi-LAT after 18
  months \protect \cite{2010ApJ...717L..71A} are also shown. \C{In the
    near future} we find it hard to detect even the brightest BM
  models, $\Ip$ and $\Kp$, where the continuum emission is dominating
  the total emission in the GeV energy range.}
 \label{fig:diff_BM}
\end{minipage}
\end{figure*}

@@

It is also interesting to compare the total contribution from the LP
model, the brightest BM model ($\Kp$), and the CR induced emission. In
Fig.~\ref{fig:flux_int} we show the integrated flux from Fornax for
our different gamma-ray models and compare it to the unresolved
integrated flux upper limit on Fornax set by \Fermi-LAT where they
averaged the flux over the energy range $0.2-100\,\gev$ assuming a
spectral index of 2. Again, the annihilation flux in the LP model is
in conflict with the upper limits by \Fermi-LAT, although only by a
factor of 30 which is \C{a factor few} less constraining than the
differential flux in the energy range $1-10$~GeV. As shown in this
figure, the LP model is dominating the entire gamma-ray energy range
up to the DM rest mass energy in this model of about 1~TeV. Finally,
we note that the flux the DM BM $\Kp$ model is larger than the
predicted emission from the CRs in the $0.1-100$~GeV energy
regime. Hence, for an experiment with a high sensitivity even for
extended sources, the prospects for detecting the $\Kp$ BM model over
the expected gamma-ray background induced by CRs looks promising in
clusters. Present day Cherenkov telescopes, however, have a trigger
region that is smaller than the size of clusters, hence it has to be
increased to several degrees to overcome problems with background
estimation. In addition, even though the projected CTA point source
sensitivity ($5\sigma$, 50h) shows the potential of this experiment in
constraining leptophilic models as well as BM models with a very large
neutralino mass $m_\chi c^2 \gtrsim 1$~TeV, we find that analysis
techniques have to be developed that enable the detection of extended
sources without too much degradation of sensitivity.

@@

\begin{figure*}
\begin{minipage}{2.0\columnwidth}
 \includegraphics[width=0.49\columnwidth]{figures/flux.cluster.Fornax.v14.0.1deg.1.6T.SubMass.SF700.IR2.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.cluster.Coma.v14.0.1deg.1.6T.SubMass.SF700.IR2.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.cluster.Virgo.v14.0.1deg.1.6T.SubMass.SF700.IR2.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.cluster.Perseus.v14.0.1deg.1.6T.SubMass.SF700.IR2.noMW.woGal.eps}
\caption{\it Comparing the gamma-ray emission from different
  clusters. We show the differential flux of clusters in the upper
  panel of each figure: Fornax (upper left), Coma (upper right), Virgo
  (lower left), and Perseus (lower right). We show CR induced emission
  (red solid), a leptophilic (LP) model that includes both final state
  radiation and IC upscattered CMB, dust and starlight (green
  dash-dotted), and the benchmark $\Kp$ model (BM) that includes
  continuum emission, and IC upscattered CMB, dust and starlight (blue
  dashed). The arrows show the spatially extended differential upper
  limits set by \Fermi-LAT in the energy ranges $0.2-1$~GeV,
  $1-10$~GeV, and $10-100$~GeV from left to right (using Gauss
  profiles for the source extension with $\sigma=(1.2,1.0,0.8)$ for
  (Virgo, Coma, Fornax) and an X-ray-inferred King profile for
  Perseus) \cite{2010ApJ...717L..71A}. We show the individual
  components for the BM model and LP model for Virgo and Perseus,
  respectively. In the lower panel of each figure we show differential
  spectral indices, $\alpha_\gamma$, where $\dd F_\gamma/\dd E_\gamma
  \sim E_\gamma^{-\alpha_\gamma}$. The flux from the clusters is
  integrated out to $\rvir$ using a point spread function of $0.1\deg$
  and includes the boost from substructures. We find that the lower
  GeV-energy regime is most constraining due to the peak sensitivity
  of \Fermi at these energies and use this regime to get upper limits
  on boost factors. \C{Note that the CR-induced emission in
    Coma is close to the upper limits set by \Fermi and will be tested
    in the upcoming years.}}
 \label{fig:clu_comp}
\end{minipage}
\end{figure*}

@@

\begin{figure*}
\begin{minipage}{2.0\columnwidth}
  \includegraphics[width=0.99\columnwidth]{figures/Fermi.comp.DM.eps}
  \caption{\it \Fermi gamma-ray flux upper limits are contrasted to
    predicted DM gamma-ray fluxes. We show the mean differential flux
    in the energy range $E_\gamma=1-10$~GeV for 32 clusters. The
    maximally extended \Fermi-LAT upper limits are shown with black
    arrows, while the point source limits are shown in grey. The
    predicted fluxes are derived from the dominant inverse
    Compton-scattering of CMB photons in a leptophilic DM model (red),
    and the continuum emission from the DM $\Kp$ benchmark model
    (blue). Assuming a boost factor due to substructures that has a
    mass spectrum extending down to an Earth mass, the expected
    leptophilic fluxes are ruled out by upper limits in 28 of the
    clusters, with the strongest constraints set by M49 and
    Fornax. \C{At the present time, we cannot constrain the benchmark
      models, although \Fermi-LAT will soon start to probe these
      models. In particular, improved modelling of extended sources as
      well as stacking of clusters appears promising in testing
      benchmark models.}}
 \label{fig14}
\end{minipage}
\end{figure*}

>>>>

In this paper, we will investigate in some more detail the potential
of several of the most promising galaxy clusters to produce an
annihilation gamma-ray yield which could be observable with present
and planned gamma-ray detectors. For convenience, we summarize in
Table~\ref{tab:flux_tab} our results for the gamma-ray flux from four
promising clusters with a high gamma-ray yield. We will in the
following derive the emission from two different DM models and a
hadronic CR model to show that Virgo has the highest gamma-ray flux
from annihilating DM. However, if the boost from Sommerfeld
enhancement (SFE) is small then the gamma-ray emission induced by CRs
is dominating most clusters with the exception of a few clusters; most
notably Fornax that shows the largest ratio of DM annihilation-to-CR
flux. In particular, we identify Fornax and Virgo as prime targets for
DM observations, Perseus for CR induced emission and we additionally
decide in favor of the well studied cluster Coma for comparison. While
studying these four clusters in great detail, we complement the
analysis by calculating the expected gamma-ray fluxes of a nearby
X-ray flux complete sample of galaxy clusters from the Rosat all-sky
survey (extended HIghest X-ray FLUx Galaxy Cluster Sample, HIFLUGCS,
\citealt{2002ApJ...567..716R,2007A&A...466..805C}).\footnote{Note that
  we have added the Virgo cluster to the sample that we nevertheless
  refer to as extended HIFLUGCS catalogue in the following.} For
previous work related to DM in clusters, see, e.g.,
\cite{2006A&A...455...21C,2009PhRvD..80b3005J,2011arXiv1104.3530S,2011ApJ...726L...6C,2010JCAP...05..025A}
and for recent work on CRs in clusters see
\cite{2010MNRAS.409..449P,2011MNRAS.410..127B,2008MNRAS.385.1211P,2009JCAP...08..002K,2010MNRAS.407.1565D}.

\begin{table*}
\begin{minipage}{2.0\columnwidth}
  \caption{Gamma-ray flux from various clusters within $\rvir$.}
\begin{tabular}{l c c c c c c c c c c}
\hline
\hline
 Cluster &
\multicolumn{3}{c}{$F_{\gamma}(>100\,\rmn{MeV})$ $[\rmn{ph}\,\rmn{cm}^{-2}\,\rmn{s}^{-1}]$:} & &
\multicolumn{3}{c}{$F_{\gamma}(>100\,\rmn{GeV})$ $[\rmn{ph}\,\rmn{cm}^{-2}\,\rmn{s}^{-1}]$:} &
$\mvir$ $^{(1)}$ & $\B_\rmn{sfe} $$^{(2)}$ & $\B_\rmn{sub} $$^{(3)}$ \\
         & DM-LP $^{(4)}$ & DM-BM-$\Kp$ $^{(5)}$ & CR-$\pi^0$ $^{(6)}$
         & & DM-LP $^{(4)}$ & DM-BM-$\Kp$ $^{(5)}$ & CR-$\pi^0$ $^{(6)}$ & $[10^{14}\,\msun]$ && \\
 \hline
 Coma & $8.2\times10^{-8}$ & $7.6\times10^{-11}$ & $4.1\times10^{-9}$
 & \,\,\,\,\, & $1.2\times10^{-11}$ & $2.8\times10^{-13}$ & $1.5\times10^{-12}$
 & $12.9$ & $530/65$ & $1290$ \\
 Perseus \,\,\,\,\,\, & $8.6\times10^{-8}$ & $7.8\times10^{-11}$ & $1.5\times10^{-8}$
 & \,\,\,\,\, & $1.2\times10^{-11}$ & $2.9\times10^{-13}$ & $5.5\times10^{-12}$
 & $8.6$ & $530/75$ & $1190$ \\
 Virgo & $1.6\times10^{-6}$ & $1.5\times10^{-9}$ & $1.5\times10^{-8}$
 & \,\,\,\,\, & $2.3\times10^{-10}$ & $5.4\times10^{-12}$ & $5.7\times10^{-12}$
 & $6.9$ & $530/80$ & $1120$ \\
 Fornax & $3.5\times10^{-7}$ & $3.2\times10^{-10}$ & $3.1\times10^{-10}$
 & \,\,\,\,\, & $5.1\times10^{-11}$ & $1.2\times10^{-12}$ & $1.1\times10^{-13}$
 & $2.4$ & $530/110$ & $890$ \\
\hline
\hline
\end{tabular}
\begin{quote}
  Notes: \\
  (1) The mass of Fornax, Coma, and Perseus are derived from \cite{2007A&A...466..805C},
  while the mass of Virgo is derived from \cite{1984ApJ...281...31T}.\\
  (2) The boost due to Sommerfeld enhancement. The first value shows the saturated
  Sommerfeld boost realized when substructures are present, the latter is the Sommerfeld
  boost without substructures (see Section II A 1).\\
  (3) The boost due to substructures relative a the smooth DM
  distribution. We integrate the emission in a cylinder with a radial
  extent of $2.5 R_{200}$ along the line-of-sight and an angular size
  corresponding to $\rvir$ (both measured from the cluster center).\\
  (4) The total gamma-ray flux from leptophilic (LP) DM where both the
  boost from substructures and Sommerfeld enhancement are
  included. Note that all these values are in conflict with upper limits from
  \Fermi (see the following sections for detail).\\
  (5) The total gamma-ray flux from the supersymmetric $\Kp$ benchmark (BM)
  model where the boost from substructures is included. See Sections II A 2 and II B 3.\\
  (6) Gamma-ray flux induced by CR protons. See Section V C.
 \label{tab:flux_tab}
  \end{quote}
\end{minipage}
\end{table*}

@@

We adopt our benchmark LP model from \cite{Finkbeiner:2010sm} where we
use a DM mass $m_\chi=1.6$~TeV and a branching ratio of
($1/4:1/4:1/2$) into ($\mu^+\mu^-:\e^+\e^-:\pi^+\pi^-$). Furthermore,
since most of the gamma-ray flux is expected to come from dense
subhalos within clusters which have a velocity dispersion close to the
velocity limit $v_\rmn{sat}$ where the SFE saturates, we use
$S=\B_\sfe(v_\rmn{sat})\approx 530$ for the cluster
halos.\footnote{Note that $S$ is strictly speaking the Sommerfeld
  enhanced annihilation cross-section, but here we have additionally
  included the order unity constraint that ensures that the relic
  density is not over-depleted at freeze-out.} However, for figures
where we neglect the substructure boost, we adopt a velocity dependent
SFE that is normalized to fit the electron and positron excess at
Earth observed by the PAMELA and \Fermi satellites,
\begin{equation}
%\B_\sfe(\sigv) = 0.266\, \frac{c}{v_\rmn{sat}}\, \frac{v_\rmn{MW}}{\sigv} =
%121 \left(\frac{\mvir}{10^{15}\,\msun}\right)^{-1/3}\,.
\B_\sfe(\sigv) = \B_\rmn{sfe,MW} \frac{v_\rmn{MW}}{\sigv} =
70 \left(\frac{\mvir}{10^{15}\,\msun}\right)^{-1/3}\,.
\label{eq:B_sfe}
\end{equation}
Here the boost factor needed to explain the PAMELA/\Fermi excess of
electrons and positrons $\B_\rmn{sfe,MW}=300$, the velocity dispersion
for the MW $v_\rmn{MW} \approx 220\,\rmn{km\,s}^{-1}$, and the
velocity dispersion $\sigv$ for clusters is derived in
\cite{2005RvMP...77..207V}. Clusters have a larger velocity dispersion
and an accordingly smaller Sommerfeld boost factor.

@@

\subsubsection{Substructures}
\label{sect:subst}
High-resolution dissipationless DM simulations of MW type halos find
substantial amount of substructures in the periphery of DM halos,
while the substructures in the center suffer from dynamical friction
and tidal effects depleting their central number densities. Since the
DM annihilation rate depends on the density squared, the resulting
flux from substructures is boosted compared to the smooth density
distribution. While there is still a discrepancy in the literature of
the exact value of the predicted boost factor of the DM annihilation
luminosity from substructures in DM halos, this inconsistency starts
to become resolved, apparently converging toward predictions at the
high end \protect \cite{2008MNRAS.391.1685S, 2008Natur.456...73S,
  2010ApJ...718..899A}. Following recent high-resolution simulations
of Galaxy-sized halos, we adopt a boost factor due to substructures of
230 for such a halo \cite{2008MNRAS.391.1685S} (which needs to be
scaled to cluster halos as we will discuss below).

The initial suggestions of a small {\em total} boost ($<R_{200}$) of
order unity by the Via Lactea simulations \cite{2008JPhCS.125a2008K}
made assumptions of computing the boost assuming that the substructure
luminosity follows that of the smooth DM distribution. Physically,
this would imply substructure to follow the smooth DM distribution and
a radially independent concentration parameter of substructures. Both
assumptions are in conflict with tidal mass loss of satellites which
is at work in simulations and cause the radial number density to be
anti-biased with respect to the host's mass density profile yielding a
significantly flatter subhalo distribution compared to the smooth DM
distribution \cite{2010ApJ...718..899A}. This increases the
substructure boost preferentially in the outer parts of DM halos with
a factor that ranges between 20 and 1000 (for dwarfs and galaxy
clusters), hence depending sensitively on the host halo mass (see
Figure~\ref{fig:radial_lum}). Recent {\it Phoenix} simulations by Gao
et al. \cite{2011arXiv1107.1916G} confirm the large enhancement due to
substructures in clusters. In fact they find a boost of 1125 for a
Coma like cluster which agrees within 10\% with our estimated boost
(see e.g. Table~\ref{tab:flux_tab}).

@@

In Fig.~\ref{fig:diff_BM} we show the differential flux from the
Fornax cluster which is one of the best clusters for indirect DM
searches due to its high DM annihilation fluxes and low CR induced
fluxes. We show the emission of four different supersymmetric BM
models and contrast it to the emission induced by CRs. Comparing this
emission to the differential gamma-ray upper limits set by \Fermi-LAT,
we find that the upper limits are not violated. The predicted DM flux
that is dominated by the continuum emission from $\Kp$ and $\Ip$
models (shown in the left panels) are about a factor few below the
upper limits, making it hard for \Fermi to probe these kind of DM
models in the near future without a significant improvement in the
analysis from e.g. stacking of clusters and improved methods for
analyzing extended sources. Furthermore, the gamma-ray signal induced
by CRs is expected to be about a factor 10 below the DM continuum flux
from the $\Kp$ and $\Ip$ BM models at 10~GeV. However, the IC emission
from upscattered CMB and SD photons is at least a factor of a few
lower for $\Kp$ and a factor 1000 lower for $\Ip,\Jp,\Js$, than the
expected flux from CRs above 100~MeV, making it very hard to
distinguish such a signal from the foreground due to the similar
spectral index. For energies below 100~MeV we expect IC emission by
primary shock accelerated electrons to be dominating
\cite{2010MNRAS.409..449P} over the supersymmetric DM induced
leptons. Hence in clusters, the IC emission from supersymmetric DM BM
models can be neglected compared to the CR pion and DM continuum
emission.

\begin{figure*}
\begin{minipage}{2.0\columnwidth}
 \includegraphics[width=0.49\columnwidth]{figures/flux.BMcompI.v14.0.1deg.1.6T.SubMass.IR2.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.BMcompJ.v14.0.1deg.1.6T.SubMass.IR2.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.BMcompK.v14.0.1deg.1.6T.SubMass.IR2.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.BMcompJs.v14.0.1deg.1.6T.SubMass.IR2.noMW.woGal.eps}
\caption{\it Comparing the differential flux from different models: we
  show the continuum emission from DM benchmark (BM) models (blue
  dashed), electrons and positrons from DM BM models that
  Compton-upscatter CMB photons (green dotted) as well as dust and
  star photons (orange dash-dotted), and CR induced gamma-ray emission
  (red solid). Each panel is associated with an individual DM BM
  model; upper left $\Ip$, lower left $\Kp$, upper right $\Jp$, and
  lower right $\Js$. The emission is calculated for the Fornax cluster
  using a point spread function of $0.1\deg$. The substructures boost
  the gamma-ray flux from IC upscattered CMB and continuum emission by
  a factor of 890 while the IC upscattered SD photons are only boosted
  by 20. Extended upper limits ($<1^\circ$) from \Fermi-LAT after 18
  months \protect \cite{2010ApJ...717L..71A} are also shown. In the
  near future we find it hard to detect even the brightest BM models,
  $\Ip$ and $\Kp$, where the continuum emission is dominating the
  total emission in the GeV energy range.}
 \label{fig:diff_BM}
\end{minipage}
\end{figure*}

@@

It is also interesting to compare the total contribution from the LP
model, the brightest BM model ($\Kp$), and the CR induced emission. In
Fig.~\ref{fig:flux_int} we show the integrated flux from Fornax for
our different gamma-ray models and compare it to the unresolved
integrated flux upper limit on Fornax set by \Fermi-LAT where they
averaged the flux over the energy range $0.2-100\,\gev$ assuming a
spectral index of 2. Again, the annihilation flux in the LP model is
in conflict with the upper limits by \Fermi-LAT, although only by a
factor of 30 which is a factor few less constraining than the
differential flux in the energy range $1-10$~GeV. As shown in this
figure, the LP model is dominating the entire gamma-ray energy range
up to the DM rest mass energy in this model of about 1~TeV. Finally,
we note that the flux the DM BM $\Kp$ model is larger than the
predicted emission from the CRs in the $0.1-100$~GeV energy
regime. Hence, for an experiment with a high sensitivity even for
extended sources, the prospects for detecting the $\Kp$ BM model over
the expected gamma-ray background induced by CRs looks promising in
clusters. Present day Cherenkov telescopes, however, have a trigger
region that is smaller than the size of clusters, hence it has to be
increased to several degrees to overcome problems with background
estimation. In addition, even though the projected CTA point source
sensitivity ($5\sigma$, 50h) shows the potential of this experiment in
constraining leptophilic models as well as BM models with a very large
neutralino mass $m_\chi c^2 \gtrsim 1$~TeV, we find that analysis
techniques have to be developed that enable the detection of extended
sources without too much degradation of sensitivity.

@@

\begin{figure*}
\begin{minipage}{2.0\columnwidth}
 \includegraphics[width=0.49\columnwidth]{figures/flux.cluster.Fornax.v14.0.1deg.1.6T.SubMass.SF700.IR2.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.cluster.Coma.v14.0.1deg.1.6T.SubMass.SF700.IR2.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.cluster.Virgo.v14.0.1deg.1.6T.SubMass.SF700.IR2.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.cluster.Perseus.v14.0.1deg.1.6T.SubMass.SF700.IR2.noMW.woGal.eps}
\caption{\it Comparing the gamma-ray emission from different
  clusters. We show the differential flux of clusters in the upper
  panel of each figure: Fornax (upper left), Coma (upper right), Virgo
  (lower left), and Perseus (lower right). We show CR induced emission
  (red solid), a leptophilic (LP) model that includes both final state
  radiation and IC upscattered CMB, dust and starlight (green
  dash-dotted), and the benchmark $\Kp$ model (BM) that includes
  continuum emission, and IC upscattered CMB, dust and starlight (blue
  dashed). The arrows show the spatially extended differential upper
  limits set by \Fermi-LAT in the energy ranges $0.2-1$~GeV,
  $1-10$~GeV, and $10-100$~GeV from left to right (using Gauss
  profiles for the source extension with $\sigma=(1.2,1.0,0.8)$ for
  (Virgo, Coma, Fornax) and an X-ray-inferred King profile for
  Perseus) \cite{2010ApJ...717L..71A}. We show the individual
  components for the BM model and LP model for Virgo and Perseus,
  respectively. In the lower panel of each figure we show differential
  spectral indices, $\alpha_\gamma$, where $\dd F_\gamma/\dd E_\gamma
  \sim E_\gamma^{-\alpha_\gamma}$. The flux from the clusters is
  integrated out to $\rvir$ using a point spread function of $0.1\deg$
  and includes the boost from substructures. We find that the lower
  GeV-energy regime is most constraining due to the peak sensitivity
  of \Fermi at these energies and use this regime to get upper limits
  on boost factors. Note that the CR-induced emission in Coma is close
  to the upper limits set by \Fermi and will be tested in the upcoming
  years.}
 \label{fig:clu_comp}
\end{minipage}
\end{figure*}

@@

\begin{figure*}
\begin{minipage}{2.0\columnwidth}
  \includegraphics[width=0.99\columnwidth]{figures/Fermi.comp.DM.eps}
  \caption{\it \Fermi gamma-ray flux upper limits are contrasted to
    predicted DM gamma-ray fluxes. We show the mean differential flux
    in the energy range $E_\gamma=1-10$~GeV for 32 clusters. The
    maximally extended \Fermi-LAT upper limits are shown with black
    arrows, while the point source limits are shown in grey. The
    predicted fluxes are derived from the dominant inverse
    Compton-scattering of CMB photons in a leptophilic DM model (red),
    and the continuum emission from the DM $\Kp$ benchmark model
    (blue). Assuming a boost factor due to substructures that has a
    mass spectrum extending down to an Earth mass, the expected
    leptophilic fluxes are ruled out by upper limits in 28 of the
    clusters, with the strongest constraints set by M49 and Fornax. At
    the present time, we cannot constrain the benchmark models,
    although \Fermi-LAT will soon start to probe these models. In
    particular, improved modelling of extended sources as well as
    stacking of clusters appears promising in testing benchmark
    models.}
 \label{fig14}
\end{minipage}
\end{figure*}
